---
title: "CytoRSuite: Visualisation Using cyto_plot"
author: "Dillon Hammill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CytoRSuite: Visualisation Using cyto_plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Overview

**CytoRSuite** contains an intuitive plotting function built on base graphics called `cyto_plot` which supports all existing flow cytometry objects including flowFrames, flowSets, GatingHierarchies and GatingSets. Some key plotting features include:

- 1-D density distributions
- 2-D scatter plots with blue-red colour scale
- restrict percentage of events to plot using `display`
- group samples prior to plotting using `group_by`
- order samples prior to plotting using `group_by`
- full back-gating support through `overlay`
- plot all supported gates including `rectangleGate`, `polygonGate`, `ellipsoidGate` and `filters` objects.
- full customisation for gates using `gate_line_type`, `gate_line_width` and `gate_line_col`.
- label plots with text and statistics using `label_text` and `label_stat`.
- position labels using `label_box_x` and `label_box_y`
- support for legend through `legend`
- stacked density distributions with `density_stack`
- density distributions normalised to mode with `density_modal`
- gated stacked density distributions with `gate` and `density_stack`
- full customisation of density distributions using `density_fill`, `density_fill_alpha`, `density_line_type`, `density_line_width` and `density_line_col`.
- add 2-D contour lines using `contour_lines`
- customise contour lines with `contour_line_type`, `contour_line_width` and `contour_line_col`
- full customisation of point using `point_col`, `point_alpha`, `point_shape` and`point_size`.
- full support for faceting with customisable grid layout through `layout`
- ability to plot in pop-up window through `popup`
- plot entire gating schemes using `cyto_plot_gating_scheme`
- plot marker expression profiles using `cyto_plot_profiles`
- works with `png()` to export high resolution images

`cyto_plot` is used internally throughout **CytoRSuite** so any cyto_plot arguments can be passed to functions which generate plots, including `gate_draw` and `gate_edit`

# 2. Demonstration

Here we aim to document some of the key features of `cyto_plot` for visualisation of flow cytometry data. To accomplish this goal we will use the `Activation` dataset supplied with **CytoRSuiteData**.

## 2.1 Preparation of Flow Cytometry Data
```{r, message = FALSE, warning = FALSE}
# Load Required Packages
library(CytoRSuite)
library(CytoRSuiteData)

# Prepare Data
fs <- Activation

# Add Marker Names
chnls <- c("Alexa Fluor 405-A","Alexa Fluor 430-A","APC-Cy7-A", "PE-A", "Alexa Fluor 488-A", "Alexa Fluor 700-A", "Alexa Fluor 647-A", "7-AAD-A") 
markers <- c("Hoechst-405", "Hoechst-430", "CD11c", "Va2", "CD8", "CD4", "CD44", "CD69")
names(markers) <- chnls
markernames(fs) <- markers

# Add fs to GatingSet
gs <- GatingSet(fs)

# Apply Compensation
gs <- compensate(gs, fs[[1]]@description$SPILL)

# Transform Fluorescent Channels
channels <- cyto_fluor_channels(gs)
trans <- estimateLogicle(gs[[1]], channels)
gs <- transform(gs, trans)

# Apply Saved Gates
gating(Activation_gatingTemplate,gs)

# Pull out T Cells Population 
TCells <- getData(gs, "T Cells")

```

## 2.2 `cyto_plot` Overview

At its most basic level, `cyto_plot` constructs either a 1-D or 2-D plot based on the number of channels supplied to the channels argument. To construct the plot in a pop-up window simply set the popup argument to TRUE. Note: for larger plots users will need to expand the plotting area to avoid figure margins errors.

If a single channel is supplied a 1-D density distribution is plotted. All density distributions are normalised to the mode by default by this can be changed by setting `density_modal` to FALSE.

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(fs[[1]], channels = "FSC-A")
```

If two channels are supplied a 2-D scatterplot with blue-red colour gradient is constructed.

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(fs[[1]], channels = c("FSC-A","SSC-A"))
```

## 2.3 `cyto_plot`: 1-Dimensional Density Distributions

To get 1-D density distributions of the data, simply supply a single channel or marker name to the channels argument. 

### 2.3.1 flowFrame

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(TCells[[1]], channels = "CD4")
```

### 2.3.2 flowSet

The density distribution for each flowFrame within the flowSet will be plotted in a separate panel by default. Notice how the x axis in the above plot has limits [0,4] this is due to flowFrame and flowSet objects not retaining the logicle transformation information. To have appropriately labelled axes users should supply a transformList object to the `axes_trans` argument as shown below. The `density_fill` argument controls the fill colour for the density distribution.

```{r, fig.width = 11, fig.height = 5, fig.align='center'}
trans <- estimateLogicle(fs[[2]], channels)
cyto_plot(TCells, channels = "CD4", axes_trans = trans, density_fill = "dodgerblue")
```

Stacking of samples is supported through the `density_stack` argument which controls the percentage overlap of the stacked distributions. The `density_fill_alpha` argument controls the degree of transparency of the fill colour.

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(TCells, channels = "CD4", density_fill = c("dodgerblue", "red"), density_stack = 0.5, density_fill_alpha = 0.5)
```

### 2.3.3 GatingHierarchy

To plot objects of class GatingHierarchy supply the name of the `parent` population to extract for plotting. Population overlays are supported for flowFrame, flowSet, list of flowFrames or list of flowSets using the `overlay` argument. For GatingHierarchies and GatingSets the name of the population to overlay is also supported. The degree of overlap for overlays can be controlled using the `density_stack` argument. For 1-D density distribution the `density_line_col` argument controls the border colour.

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(gs[[1]], parent = "T Cells", channels = "CD4", overlay = "CD4 T Cells", density_fill = c("green","blue"), density_line_col = c("black","red"))
``` 

### 2.3.4 GatingSet

To change the border line type user can supply an integer [0,4] to the `density_line_type` argument. The line width can be controlled in a similar fashion using the `density_line_width` argument. By default the title above each plot will be the name of the sample but this can be changed by supplying specific names to the `title` argument.

```{r, fig.width = 11, fig.height = 5, fig.align='center'}
cyto_plot(gs, parent = "T Cells", channels = "CD4", density_line_type = 2, density_line_width = 4, title = c("Unactivated","Activated"))
```

## 2.4 `cyto_plot`: 2-Dimensional Scatter Plots

To get 2-D density scatter plots of the data, simply supply two channels or marker names to the **channels** argument of cyto_plot. 

### 2.4.1 flowFrame

By default 2-D scatter plots will use a blue-red colour scale for points, this can be modified using the  `point_col` argument. To modify axes labels simply supply the desired labels to the `xlab` and `ylab` arguments. The axes limits can be modified by supplying the lower and upper limits to the `xlim` and `ylim` arguments.

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(TCells[[1]], channels = c("CD4","CD8"), axes_trans = trans, xlab = "CD4", ylab = "CD8", xlim = c(-0.5,3.5), ylim = c(-0.5,4))
```

### 2.4.2 flowSet

Contour lines are also supported for 2-D scatter plots, simply supply the number of contours levels to the `contour_lines` argument. Users can use the `contour_line_col` and `contour_line_width` arguments to modify the colour and line width of the contour lines respectively.

```{r, fig.width = 11, fig.height = 5, fig.align='center'}
cyto_plot(TCells, channels = c("CD4","CD8"), axes_trans = trans, contour_lines = 15)
```

### 2.4.3 GatingHierarchy

`cyto_plot` also has full support for plotting gates, simply supply the gate object to the `gate` argument or for GatingHierarchies and GatingSet simply supply the name of the gated population to the `alias` argument. Users can use the the `gate_line_col`, `gate_line_type` and `gate_line_width` arguments to modify the gate colour, line type or line width.

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
cyto_plot(gs[[1]], parent = "T Cells", alias = c("CD4 T Cells","CD8 T Cells"), channels = c("CD4","CD8"))
```

### 2.4.4 GatingSet

Overlays are also supported in 2-D scatterplots, simply supply the name of the gated population to the **overlay**. Overlay also accepts the following objects: flowFrames, flowSets, list of flowFrames or list of flowSets which allow overlay of any data. The colour of the overlay is also controlled by the `point_col` argument, simply supply a colour for the base plot and a colour for each overlay. Here we will set the colour of the base plot to "grey" and the overlay colours to "purple" and "blue". To include a legend in the plot users can set the `legend` argument to TRUE. To show the layering used within cyto_plot we have added some contours as well.

```{r, fig.width = 12, fig.height = 5, fig.align='center'}
cyto_plot(gs, parent = "T Cells", alias = c("CD4 T Cells","CD8 T Cells"), channels = c("CD4","CD8"), contour_lines = 15, overlay = c("CD4 T Cells","CD8 T Cells"), point_col = c("grey","purple", "blue"), point_alpha = 0.5, legend = TRUE)
```

# 3. Export High Resolution Images

```{r, eval = FALSE}
png("Activation.png", res = 600, height = 5, width = 13, units = "in")

cyto_plot(gs, parent = "T Cells", alias = c("CD4 T Cells","CD8 T Cells"), channels = c("CD4","CD8"), contour_lines = 15, overlay = c("CD4 T Cells","CD8 T Cells"), point_col = c("grey","purple", "blue"), point_alpha = 0.5, legend = TRUE)

dev.off()
```

# 4. More information

For more information on these visualisation functions refer to the documentation for these functions in the [Reference](https://dillonhammill.github.io/CytoRSuite/reference/index.html).
