---
title: "CytoRSuite: Manual and Automated Gating"
author: "Dillon Hammill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CytoRSuite: Manual and Automated Gating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Overview

**CytoRSuite** is built on and shares the same gatingTemplate as the **openCyto** package. This means that users can take advantage of both the manual gating in **CytoRSuite** and the automated gating in **openCyto** for their analyses. Furthermore, **CytoRSuite** users can take advantage of some of the other features of the **openCyto** gatingTemplate, such as boolean and reference gates. More information on the **openCyto** gatingTemplate can be found [here](http://bioconductor.org/packages/release/bioc/vignettes/openCyto/inst/doc/HowToWriteCSVTemplate.html).

# 2. Demonstration

Here we aim to demonstrate the use of both manual and automated gating approaches using **CytoRSuite** and **openCyto**. For demonstration purposes we will use the `Activation` dataset shipped with **CytoRSuiteData**.

## 2.1 Preparation of Activation Dataset for Gating

```{r, message = FALSE, warning = FALSE}
library(CytoRSuite) # openCyto is imported with CytoRSuite
library(CytoRSuiteData)

# Assign Activation flowSet to fs
fs <- Activation

# Add fs to a GatingSet
gs <- GatingSet(fs)

# Apply compensation
gs <- compensate(gs, fs[[1]]@description$SPILL)

# Transform fluorescent channels
trans <- estimateLogicle(gs[[4]], cyto_fluor_channels(gs))
gs <- transform(gs, trans)
```

## 2.2 Manually Gate `Cells` 

```{r, eval = FALSE}
# gate_draw returns gatingTemplate entry
template <- gate_draw(gs,
                      parent = "root",
                      alias = "Cells",
                      channels = c("FSC-A","SSC-A"))
```

## 2.3 Manually Gate `Single Cells`

```{r, eval = FALSE}
template <- rbind(template, gate_draw(gs,
                                      parent = "Cells",
                                      alias = "Single Cells",
                                      channels = c("FSC-A","FSC-H")))
```

## 2.4 Manually Gate `Live Cells`

```{r, eval = FALSE}
template <- rbind(template, gate_draw(gs,
                                      parent = "Single Cells",
                                      alias = "Live Cells",
                                      channels = c("Alexa Fluor 405-A",
                                                   "Alexa Fluor 430-A")))
```

## 2.5 Automatically Gate `T Cells`

```{r, eval = FALSE}
template <- rbind(template, app_pop(gs,
                                    alias = "T Cells",
                                    pop = "+",
                                    parent = "Live Cells",
                                    dims = "APC-Cy7-A,PE-A",
                                    gating_method = "flowClust",
                                    gating_args = "K=3,target=c(0,3)"))
```

## 2.6 Automatically Gate `CD4 T Cells`

```{r, eval = FALSE}
template <- rbind(template, app_pop(gs,
                                    alias = "CD4 T Cells",
                                    pop = "+",
                                    parent = "T Cells",
                                    dims = "Alexa Fluor 700-A,Alexa Fluor 488-A",
                                    gating_method = "flowClust",
                                    gating_args = "K=3,target=c(3,0)"))
```

## 2.7 Automatically Gate `CD8 T Cells`

```{r, eval = FALSE}
template <- rbind(template, app_pop(gs,
                                    alias = "CD8 T Cells",
                                    pop = "+",
                                    parent = "T Cells",
                                    dims = "Alexa Fluor 700-A,Alexa Fluor 488-A",
                                    gating_method = "flowClust",
                                    gating_args = "K=3,target=c(0,3)"))
```

## 2.8 Manually Gate `CD69+ CD4 T Cells`

```{r, eval = FALSE}
template <- rbind(template, gate_draw(gs,
                                      parent = "CD4 T Cells",
                                      alias = "CD69+ CD4 T Cells",
                                      channels = c("Alexa Fluor 647-A",
                                                   "7-AAD-A"),
                                      type = "interval",
                                      axis = "y"))
```

## 2.9 Manually Gate `CD69+ CD8 T Cells`

```{r, eval = FALSE}
template <- rbind(template, gate_draw(gs,
                                      parent = "CD8 T Cells",
                                      alias = "CD69+ CD8 T Cells",
                                      channels = c("Alexa Fluor 647-A",
                                                   "7-AAD-A"),
                                      type = "interval",
                                      axis = "y"))
```

## 2.10 Save gatingTemplate for Future Use

```{r, eval = FALSE}
write.csv(template, "gatingTemplate.csv", row.names = FALSE)
```

## 2.11 Applying a Saved gatingTemplate

```{r, eval = FALSE}
# Assign Activation flowSet to fs
fs <- Activation

# Add fs to a GatingSet
gs <- GatingSet(fs)

# Apply compensation
gs <- compensate(gs, fs[[1]]@description$SPILL)

# Transform fluorescent channels
trans <- estimateLogicle(gs[[4]], cyto_fluor_channels(gs))
gs <- transform(gs, trans)

# Apply gatingTemplate
gt <- gatingTemplate("gatingTemplate.csv")
gating(gt,gs)

# Visualise Gating Scheme
cyto_plot_gating_scheme(gs[[4]])
```

# 3. More Information

For more information on manual gating using **CytoRSuite** refer to [Gating Functions](https://dillonhammill.github.io/CytoRSuite/articles/CytoRSuite-Gating-Functions.html) and [Gate Manipulation](https://dillonhammill.github.io/CytoRSuite/articles/CytoRSuite-Gate-Manipulation.html) vignettes. For more information on automated gating using **openCyto** refer to this [Gating Methods](https://www.bioconductor.org/packages/devel/bioc/vignettes/openCyto/inst/doc/HowToAutoGating.html) vignette.
